Links for original code sources are indicated when needed.

This code allows analysis of transcriptomics data (a table of gene counts per sample), resulting in a list of differentially expressed genes (DEGs). From these lists, volcano plots and Euler diagrams were created.

### DATA PREPARATION ###

```{r}
setwd("C:/Users/harri/Desktop/vario-transcripto")
```


Install packages
```{r}
#install.packages("BiocManager")
#BiocManager::install("DESeq2")
```

Load the library
```{r}
library(DESeq2)
```

Import data and sample description file (.csv)
```{r}
library(readr)

merged_gene_abundance <- read.delim("merged_gene_abundance.tsv", row.names=1) #if doesn't work From Text (base) > change row names to First column
sample_counts <- merged_gene_abundance

info_table <- read_delim("info-table-vario.csv", delim = ";", escape_double = FALSE, trim_ws = TRUE)
sample_info <- info_table

```

Convert the column from characters to factors (required by DESeq)
```{r}
sample_info$Type <- factor(sample_info$Type)
```

Set up the dds object
```{r}
sample_dds <- DESeqDataSetFromMatrix(countData = sample_counts,
                                     colData = sample_info,
                                     design = ~Type
                                     )

```


### RUN DESEQ ###

```{r}
sample_dds <- DESeq(sample_dds)
```

Quality control: Principle Components Analysis w/ regularized log transformation (rlog)

```{r}
rld <- rlog(sample_dds, blind=TRUE)
plotPCA(rld, intgroup="Type") 
```


Check result. Differential expression is comparing between two conditions as described in InfoTable. (miRNA VS SCRB)

Start by comparing at 20 min
```{r}
sample_res <- results(sample_dds, contrast = c("Type", "mir-20", "scrb-20")) #last charact vect = base level
summary(sample_res)
```

Save the results to a table
```{r}
res <- data.frame(sample_res)
res_20 <- summary(res)

library(readxl)
annotations <- read_excel("C:/Users/harri/Desktop/resultats_bruts_transcripto/variovorax/export/annotations.xlsx")
res_gene <- merge(x=res, y=annotations, by.x=0, by.y="gene_id")

write.csv(res, "DESeq-results-20min-vario.csv")
write.table(res, "DESeq-results-20min-vario.txt")
```

Filtering to find significant genes using FDR cutoff of 0.05
```{r}
padj.cutoff <- 0.05 # False Discovery Rate cutoff
significant_results <- res_gene[which(res_gene$padj < padj.cutoff),]
```

Save results using customized file_name
```{r}
write.table(significant_results, "sign-pajd-20min-vario-bis.txt")
write.csv(significant_results, "sign-pajd-20min-vario-bis.csv")

```


Now compare at 120min

```{r}
sample_res <- results(sample_dds, contrast = c("Type", "mir-120", "scrb-120")) 
summary(sample_res)
```

Save the results to a table
```{r}
res <- data.frame(sample_res)
res_120 <- summary(res)

annotations <- read_excel("~/Résultats_bruts_transcripto/variovorax/export/annotations.xlsx")
res_gene <- merge(x=res, y=annotations, by.x=0, by.y="gene_id")


write.csv(res, "DESeq-results-120min-test.csv")
write.table(res, "DESeq-results-120min-test.txt")
```

Filtering to find significant genes using FDR cutoff of 0.05
```{r}
padj.cutoff <- 0.05 # False Discovery Rate cutoff
significant_results <- res_gene[which(res_gene$padj < padj.cutoff),]
```

Save results using customized file_name
```{r}
write.table(significant_results, "sign-pajd-120min-vario.txt")
```

### Make Euler Diagram to show common differentially expressed genes (DEGs) between conditions ###

```{r}
#install.packages("eulerr")
library(eulerr)
```

Select the DEGs result files for comparison

```{r}
File20 <- file.choose() #Select the 1st DEGs result, mir-20 vs scrb-20
File120 <- file.choose() #Select the 1st DEGs result, mir-120 vs scrb-120

DEGs_20 <- read.csv(File20, header = TRUE)
DEGs_120 <- read.csv(File120, header = TRUE)

DEGs_20 <- read.csv("C:/Users/harri/Desktop/vario-transcripto/results/DESeq-results-20min-vario.txt", sep="")
DEGs_120 <- read.csv("C:/Users/harri/Desktop/vario-transcripto/results/DESeq-results-120min-test.txt", sep="")

```


Define a function to get significant DEGs
```{r}
sig_DEGs <- function(DEGs){
  df <- data.frame(DEGs)
  sig_gene_df = subset(df, padj < 0.05) 
  sig_gene_df = subset(sig_gene_df, baseMean > 10)
  sig_gene_df = subset(sig_gene_df, abs(log2FoldChange) > 1.5)
  return(sig_gene_df$X)
}

sig_DEGs <- function(DEGs){
  df <- data.frame(DEGs)
  sig_gene_df = subset(df, padj < 0.05) 
  return(sig_gene_df$X)
}
```

Get significant DEGs
```{r}
A <- sig_DEGs(DEGs_20)
B <- sig_DEGs(DEGs_120)
```


Get the intersection and difference of A and B
```{r}
AnotB <-setdiff(A, B)
BnotA <-setdiff(B,A)
AandB <-intersect(A, B)

eulerplot <-euler(c("20min"=length(AnotB), 
                    "120min"=length(BnotA), 
                    "20min&120min"=length(AandB))
                  )
```


# https://rdrr.io/cran/eulerr/man/plot.euler.html
# https://cran.r-project.org/web/packages/eulerr/vignettes/gallery.html
```{r}
plot(eulerplot, 
     main = "Common DEGs",
     quantities = list(type = "counts"),
     edges = c("blue", "red"),
     fills = c("white", "white"),
     lwd = 4,
     legend = list(side = "right"))
```

 ### Volcano plots ###
```{r}
library(ggplot2)
library(ggrepel)
library(dplyr)
```


Data loading and pre-processing
Select the DEGs result files for comparison
```{r}
File <- file.choose() 
DEGs <- read.csv(File, header = TRUE)

res <- data.frame(DEGs)
res$Symbol <- res$X
```

Label genes on volcano plot

Set cutoff for log2FoldChange for label
```{r}
FC_neg_cutoff = -1.5 # Genes with log2FoldChange less than this value will be labeled
FC_pos_cutoff = 1.5 # Genes with log2FoldChange more than this value will be labeled
title = "Volcano Plot DEGs 120min"
```

Label a certain sets of genes, change gene symbols based on need, in makeVolcanoPlot function
```{r}
makeVolcanoPlot <- function(res,FC_neg_cutoff, FC_pos_cutoff, title, geneList){
  # Set threshold to have different color for different sets of genes
  res <- res %>% mutate(threshold = ifelse(padj<=0.050 & abs(log2FoldChange)>=1.5,"A", ifelse(padj>0.05 & abs(log2FoldChange)<1.5, "B", "C")))
    
  ## Two ways to label desired genes (run only one of them)
  # 1. Only label genes based on the cutoffs
  res$Label <- ifelse(res$log2FoldChange < FC_neg_cutoff & res$padj < 0.05 | res$log2FoldChange > FC_pos_cutoff & res$padj < 0.05, res$Symbol, "")
  
  # 2. Only label genes based on pre-defined gene list (Run either cutoff or pre-defined gene list)
  #res$Label <- ifelse(res$Symbol %in% geneList, res$Symbol, "")
  
  # Define plotting function, adjust parameters based on need
  ggplot(res, aes(log2FoldChange, -log10(padj), label = Label))+
    # Set color for different data points
    geom_point(aes(colour = threshold), size =2, alpha = 0.4) +
    scale_colour_manual(values = c("A"="red", "B"="grey", "C"="darkgreen"))+
    # Avoid text overlap
    geom_text_repel(box.padding = 0.5, max.overlaps = Inf)+
    # Set theme of the plot
    theme_bw()+
    xlim(c(-5,5))+ # x-axis range, adjust based on need
    ylim(c(0,5))+ # y-axis range,adjust based on need
    geom_vline(xintercept = c(-1.5, 1.5), lty=2,col="black",lwd=0.6)+ # Draw horizontal dash line
    geom_hline(yintercept = 1.301, lty=2,col="black",lwd=0.6)+ # Draw vertical dash line
    xlab(bquote(~Log[2]~ 'fold change'))+ # x-axis label
    ylab(bquote(~-Log[10]~ 'padj')) + # y-axis label
    theme(axis.text = element_text(color = "black", size = 10))+ # Font type and size
    ggtitle(title) # Add title
}

makeVolcanoPlot(res,FC_neg_cutoff, FC_pos_cutoff, title, geneList)
```



 Preview of the volcano plot, helps define some parameters quickly
Make an enhanced version of the volcano plot (https://bioconductor.org/packages/release/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html)
```{r}
library(EnhancedVolcano)
#BiocManager::install('EnhancedVolcano')
EnhancedVolcano(res, lab = res$Symbol,
                x = 'log2FoldChange', y = 'padj',
                pCutoff = 0.05, FCcutoff = 1.5, 
                pointSize =3, labSize = 4)
  
```







